A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-12-02 main.tex
4.2 edits - still in progress

Both PC-lasso and LMM-lasso rely on matrix decompositions of $\bX$ and the realized relationship matrix (RRM), $\bK = p^{-1} \bX \bX^\T$, which describes the pair-wise genetic similarity among subjects \citep{hayes2009increased}. Using the singular value decomposition (SVD) of the standardized genotype matrix, we can derive the spectral decomposition of the RRM:

@@

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

%------------------------------%
\subsection{\anna{Lasso and PC-lasso select more population-informative SNPs than LMM-Lasso}}
%------------------------------%

 \begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/popInfSNPs_n=200.png}
    \caption{Causal and non-causal SNP selection in the presence of varying levels of population informative SNPs with a coarse 2 Independent Subpopulation structure, dichotomous-discordant effects, $\eta = 0.5$, and $\xi = 0.8$.}
    \label{fig:big_small_gamma}
\end{figure}

>>>>

Both PC-lasso and LMM-lasso rely on matrix decompositions of $\bX$ and the realized relationship matrix (RRM), $\bK = p^{-1} \bX \bX^\T$, which describes the pair-wise genetic similarity among subjects \citep{patterson2006population, hayes2009increased}. Using the singular value decomposition (SVD) of the standardized genotype matrix, we can derive the spectral decomposition of the RRM:

@@

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) \anna{technically this is true sign rate - only estimates with the correct sign are counted. Explain this? Calculate overall selection instead?} for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

%------------------------------%
\subsection{\anna{PC-lasso and LMM-lasso reduce bias towards selecting population-informative SNPs}}
%------------------------------%

In the previous section we saw how in the presence of confounding, LMM-lasso and PC-lasso outperform the ordinary lasso in terms of selection accuracy. Additionally, as we described in Section \ref{sec:example}, subpopulation differences in allele frequencies are known to have the potential to lead to spurious associations. As such it is of interest to characterize these methods' selection accuracy in terms of how robust they are to allele frequency differences.

In order to assess the behavior of these methods in this context, we conducted a simulation using the parameter settings presented in Figure \ref{fig:mse} and Table \ref{tab:var_sel} and a modified version of the Empirical data. In order to quantify the absolute difference in subpopulation-specific allele frequencies in an straightforward, interpretable manner, we limited the Empirical data to that of the white and African American subpopulations. Figure \ref{fig:pop_inf} shows the relationship between a given SNP's false positive rate (FPR) and its absolute difference in standardized allele frequency between the two supopulations.

 \begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/pop_inf_snps_loess.png}
    \caption{\anna{Non-causal SNP selection / SNP FPR (?)} in the presence of varying levels of population informative SNPs and environmental confounding for two Empirical subpopulations, dichotomous-discordant effects, $\eta = 0.5$.}
    \label{fig:pop_inf}
\end{figure}

\anna{maybe no table ? If table, make more like table 1}
\begin{table}[H]
\centering
\begin{tabular}{rrlr}
  \hline
eta & xi & Method & $r_s$ \\ 
  \hline
0.5 & 0.0 & lasso & 0.34 \\ 
  0.5 & 0.0 & PC-lasso & -0.92 \\ 
  0.5 & 0.0 & LMM-lasso & -0.89 \\ 
  0.5 & 0.8 & lasso & 0.78 \\ 
  0.5 & 0.8 & PC-lasso & -0.82 \\ 
  0.5 & 0.8 & LMM-lasso & -0.67 \\ 
   \hline
\end{tabular}
\caption{Spearman's rank correlation coefficient between \anna{non-causal SNP selection / SNP FPR (?)} and absolute difference in standardized allele frequencies across methods.}
\label{tab:pop_inf}
\end{table}

\anna{Another way to quantify this...(?)}
\begin{table}[H]
\centering
\begin{tabular}{rlrrr}
  \hline
xi & chunks & lasso & lasso\_pca10 & plmm\_lasso \\ 
  \hline
0.0 & 1 & 0.108 & 0.187 & 0.135 \\ 
  0.0 & 2 & 0.394 & 0.132 & 0.221 \\ 
  0.0 & 3 & 0.214 & 0.049 & 0.071 \\ 
  0.8 & 1 & 0.005 & 0.437 & 0.161 \\ 
  0.8 & 2 & 0.504 & 0.188 & 0.348 \\ 
  0.8 & 3 & 0.966 & 0.084 & 0.157 \\ 
   \hline
\end{tabular}
\caption{Max FPR for each method, $\xi$, and level of difference (low, med, high)}
\end{table}

In the absence of environmental confounding, the lasso demonstrates a moderately positive correlation with allele frequency differences ($r_s = 0.34$), while the PC-lasso and LMM-lasso methods are negatively correlated with allele frequency differences.

Overall, however, the FPR does not exceed 0.39, 0.19, and 0.22 for the lasso, PC-lasso, and LMM-lasso methods, respectively.  
%\anna{We attribute these associations to the presence of population structure (in other words, correlated observations) and thus artificially shrunken errors in the case of the lasso. For PC-lasso and LMM-lasso, the absence of a non-genetic mechanism by which population structure directly affects the outcome, results in a slight over-correction for both, resulting in their moderately negative correlations. }

Table \ref{tab:pop_inf} displays the Spearman's rank correlation coefficient for each of these scenarios.

In the presence of confounding, however, lasso's correlation with allele frequency increases, and particularly struggles to correctly select the most population-informative SNPs. PC-lasso and LMM-lasso are able to remain more robust to allele frequency differences, yet remain somewhat negatively correlated with allele frequency differences. This is is advantageous in preventing dramatically high FPRs in the presence of highly population-informative SNPs.

In both scenarios, LMM-lasso is more robust to allele frequency differences than PC-lasso due to the presence of \anna{apparent cryptic relatedness (?)} \anna{max absolute kinship? histograms of off-diag kinship?} in the Empirical data. This echoes the superior TPRs of LMM-lasso compared to that of the PC-lasso in the case of the Empirical data in Table \ref{tab:var_sel}.
